Parsing Postfix log files

### 2009-01-16 doc/logparser-architecture.tex
Some minor changes based on the SGAI paper.
Converted from article to report, and sections to chapters. Some cleanups. @ doesn't need escaping in glossary entries with new glossaries package.
Lots of minor improvements and cleanups to the State of the Art Review,     including running diction on it. Checked that I've written 'log files' and 'log lines' everywhere, instead of     'files' or 'lines', and added a make target to help with it.  This is     supported by paragraphs-to-lines, which will hopefully find other uses.  I     renamed a couple of included files too, to avoid false positives.

\section{Parser Architecture}

@@

\subsection{Parser architecture and design}

@@

\begin{table}[ht]
    \caption{Similarities between ATN and this architecture}
    \empty{}\label{Similarities between ATN and this architecture}
    \begin{tabular}[]{lll}
        \tableline{}
        \gls{ATN}          & Architecture  & Similarity                \\
        \tableline{}
        Networks        & Algorithm     & Determines the sequence 
                                        of transitions              \\
                        &               & or actions that 
                                        constitutes a valid  input. \\
        Transitions     & Actions       & Save data and impose
                                        conditions the              \\
                        &               & input must meet to be
                                        accepted as valid.          \\
        Abbreviations   & Rules         & Responsible for 
                                        classifying input.          \\
        \tableline{}
    \end{tabular}
\end{table}

\subsection{Rules}

@@

\subsubsection{Rule characteristics}

@@

\subsubsection{Overlapping rules}

@@

    \item Compare the results of parsing using sorted, shuffled, and
        reversed rules (described in \sectionref{rule ordering for
        efficiency}).  Parse several log files using optimal ordering, then
        dump a textual representation of the rules, connections, and
        results tables.  Repeat with shuffled and reversed ordering,
        starting with a fresh database.  If there are no overlapping rules
        the tables from each run will be identical; differences indicate
        overlapping rules.  The rules that overlap can be determined by
        examining the differences in the tables: each result contains a
        reference to the rule which created it, if the references differ
        between runs the two rules referenced in the differing results
        overlap.  Unfortunately this method cannot prove the absence of
        overlapping rules; it can detect overlapping rules, but only if
        there are log lines in the input files that match more than one
        rule.

@@

\subsection{Framework}

XXX PULL CONTENT FROM THE PAPER FOR HERE\@; IT'S BETTER WRITTEN\@.

@@

\subsection{Actions}

>>>>

\chapter{Parser Architecture}

@@

\section{Parser architecture and design}

@@

\begin{table}[ht]
    \caption{Similarities between ATN and this architecture}
    \empty{}\label{Similarities between ATN and this architecture}
    \begin{tabular}[]{lll}
        \tabletopline{}%
        \gls{ATN}          & Architecture  & Similarity                \\
        \tablemiddleline{}%
        Networks        & Algorithm     & Determines the sequence 
                                        of transitions              \\
                        &               & or actions that 
                                        constitutes a valid  input. \\
        Transitions     & Actions       & Save data and impose
                                        conditions the              \\
                        &               & input must meet to be
                                        accepted as valid.          \\
        Abbreviations   & Rules         & Responsible for 
                                        classifying input.          \\
        \tablebottomline{}%
    \end{tabular}
\end{table}

\section{Rules}

@@

\subsection{Rule characteristics}

@@

\subsection{Overlapping rules}

@@

    \item Compare the results of parsing using sorted, shuffled, and
        reversed rules (described in \sectionref{rule ordering for
        efficiency}).  Parse several log files using optimal ordering, then
        dump a textual representation of the rules, connections, and
        results tables.  Repeat with shuffled and reversed ordering,
        starting with a fresh database.  If there are no overlapping rules
        the tables from each run will be identical; differences indicate
        overlapping rules.  The rules that overlap can be determined by
        examining the differences in the tables: each result contains a
        reference to the rule which created it, if the references differ
        between runs the two rules referenced in the differing results
        overlap.  Unfortunately this method cannot prove the absence of
        overlapping rules; it can detect overlapping rules, but only if
        there are log lines in the log files that are matched by more than
        one rule.

@@

\section{Framework}

XXX PULL CONTENT FROM THE PAPER FOR HERE\@; IT IS BETTER WRITTEN\@.

@@

\section{Actions}

### 2009-01-16 doc/logparser-background.tex
Some minor changes based on the SGAI paper.
Converted from article to report, and sections to chapters. Some cleanups. @ doesn't need escaping in glossary entries with new glossaries package.
Lots of minor improvements and cleanups to the State of the Art Review,     including running diction on it. Checked that I've written 'log files' and 'log lines' everywhere, instead of     'files' or 'lines', and added a make target to help with it.  This is     supported by paragraphs-to-lines, which will hopefully find other uses.  I     renamed a couple of included files too, to avoid false positives.

\section{Background}

@@

\subsection{Motivation}

@@

\begin{table}[ht]
    \caption{Sample output from SQL query showing percentages}
    \empty{}\label{Sample output from SQL query showing
    percentages}
    \begin{tabular}[]{lrr}
        \tableline{}
        Restriction & Number of hits & Percentage of hits \\
        \tableline{}%
        \input{build/include-sample-sql-output}
        \tableline{}
    \end{tabular}
\end{table}

@@

\subsection{Database as Application Programming Interface}

@@

The database populated by this program provides a simple interface to
Postfix log files.  Although the interface is a database schema rather than
a set of functions provided by shared code, it is in effect quite similar
to any other \gls{API}: it provides a stable interface for users of the
interface, allowing the implementation to be changed without adverse
effects.  As long as the \gls{API} is maintained, the parser can be
improved to handle additional complications; support can be added for
earlier or later releases of Postfix; bugs can be fixed or limitations
removed.

@@

            \item New columns can be added to the tables used by the
                program, with sufficient DEFAULT clauses or a clever
                TRIGGER or two.\footnote{Please refer to an \gls{SQL} guide
                for explanations of these terms, e.g.\
                \url{http://philip.greenspun.com/sql/}.}

            \item A VIEW gives a custom arrangement of data with very
                little effort.

@@

\subsection{Simple Mail Transfer Protocol}

@@

The \acrlong{SMTP}, originally defined in \gls{RFC}~821~\cite{RFC821} and
updated in \gls{RFC}~2821~\cite{RFC2821}, is used for transferring mail
between the sending and receiving \gls{MTA}\@.  It is a simple, human
readable, plain text protocol, making it quite easy to test and debug
problems with it.  Despite the simplicity of the protocol many virus and
spam sending programs fail to implement it properly, so requiring strict
adherence to the protocol specification is beneficial in protecting against
spam and viruses.\footnote{\label{footnote:rfc760}Originally all mail
servers adhered to the principle of \textit{Be liberal in what you accept,
and conservative in what you send\/} from \gls{RFC}~760~\cite{rfc760}, but
unfortunately that principle was written in a friendlier time.  Given the
deluge of spam that mail servers are subjected to daily, a more appropriate
maxim could be: \textit{Require strict adherence to \gls{RFC}~2821;
implement the strongest restrictions you can; relax the restrictions and
adherence only when legitimate mail is impeded.\/}  it is not as friendly,
nor as catchy, but it more accurately reflects the current situation.} A
typical \gls{SMTP} conversation resembles the following (the lines starting
with a three digit number are sent by the server, all other lines are sent
by the client):

@@

\subsection{Postfix}

@@

Postfix is a \acrlong{MTA}~(\gls{MTA}) with the following design aims (in
order of importance): security, flexibility of configuration, scalability,
and high performance.  It features extensive optional anti-spam
restrictions, allowing an administrator to deploy those restrictions which
they judge suitable for their site's needs, rather than a fixed set chosen
by Postfix's author.  These restrictions can be selectively applied,
combined, and bypassed on a per-client, per-recipient, or per-sender basis,
allowing varying levels of stricture and permissiveness.  Postfix leverages
simple lookup tables to support arbitrarily complicated user-defined
sequences of restrictions and exceptions; policy servers
(\sectionref{policy servers}) provide a simple way to write new
restrictions.  Administrators can also supply their own rejection messages
to make it clear to senders exactly why their mail was rejected.
Unfortunately this flexibility has a cost: complexity in the log files
generated.  Although it is easy to use standard Unix text processing
utilities to determine the fate of an individual email, following the
journey an email takes through Postfix can be quite difficult.  For most
mails the journey is simple and brief, but the remaining minority can have
quite complex journeys (see \sectionref{complications} for details).

@@

\subsubsection{Mixing and matching Postfix restrictions}

@@

\subsubsection{Policy servers}

@@

    \item Greylisting~\cite{greylisting} is a technique that temporarily
        rejects a delivery attempt when the triple of \newline{}
        \tab{}\texttt{(sender address, recipient address, remote \gls{IP}
        address)}\newline{} has not been seen before; on second and
        subsequent delivery attempts from that triple the mail will be
        accepted.  This blocks spam because maintaining a list of failed
        addresses and retrying after a temporary failure is uneconomical
        for a spam sender, but a legitimate mail server must retry
        deliveries that temporarily failed.  Sadly spam senders are using
        increasingly complex and well written programs to distribute spam,
        frequently using an ISP provided \gls{SMTP} server from a
        compromised machine on the ISP's network.  Greylisting will slowly
        become less useful, but it does block a large percentage of spam
        mail at the moment; the most effective restrictions from the
        \numberOFlogFILES{} log files used in testing the parser are shown
        in \tableref{Summary of rejections}.  Greylisting is obviously
        worth using, at least at the moment, particularly when you consider
        Greylisting's position as the final restriction that a mail must
        overcome:\footnote{Greylisting is the final restriction a mail must
        overcome in the configuration used on the mail server the log files
        were obtained from; an administrator is free to use Greylisting
        wherever in the restriction list they feel is most appropriate for
        their mail system.} Greylisting only takes effect for mails that
        have passed every other restriction.  Some problems may be
        encountered when using greylisting: some servers fail to retry;
        legitimate mail may be delayed, particularly when coming from a
        pool of servers.

>>>>

\chapter{Background}

@@

\section{Motivation}

@@

\begin{table}[ht]
    \caption{Sample output from SQL query showing percentages}
    \empty{}\label{Sample output from SQL query showing
    percentages}
    \begin{tabular}[]{lrr}
        \tabletopline{}%
        Restriction & Number of hits & Percentage of hits \\
        \tablemiddleline{}%
        \input{build/include-sample-sql-output}
        \tablebottomline{}%
    \end{tabular}
\end{table}

@@

\section{Using a Database as an Application Programming Interface}

@@

The database populated by this program provides a simple interface to
Postfix log files.  Although the interface is a database schema rather than
a set of functions provided by shared code, it performs the same function
as any other \gls{API}: it provides a stable interface for users of the
interface, allowing the implementation to be changed without adverse
effects.  If as the \gls{API} is maintained, the parser can be improved to
handle additional complications; support can be added for earlier or later
releases of Postfix; bugs can be fixed or limitations removed.

@@

            \item New columns can be added to the tables used by the
                program, using DEFAULT clauses or TRIGGERS to populate
                them.\footnote{Please refer to an \gls{SQL} guide for
                explanations of these terms, e.g.\
                \url{http://philip.greenspun.com/sql/}.}

            \item A VIEW gives a custom arrangement of data with minimal
                effort.

@@

\section{Simple Mail Transfer Protocol}

@@

The \acrlong{SMTP}, originally defined in \gls{RFC}~821~\cite{RFC821} and
updated in \gls{RFC}~2821~\cite{RFC2821}, is used for transferring mail
between the sending and receiving \gls{MTA}\@.  It is a simple, human
readable, plain text protocol, simplifying testing and debugging problems
with it.  Despite the simplicity of the protocol many virus and spam
sending programs fail to implement it properly, so requiring strict
adherence to the protocol specification is beneficial in protecting against
spam and viruses.\footnote{\label{footnote:rfc760}Originally all mail
servers adhered to the principle of \textit{Be liberal in what you accept,
and conservative in what you send\/} from \gls{RFC}~760~\cite{rfc760}, but
unfortunately that principle was written in a friendlier time.  Given the
deluge of spam that mail servers are subjected to daily, a more appropriate
maxim could be: \textit{Require strict adherence to \gls{RFC}~2821;
implement the strongest restrictions you can; relax the restrictions and
adherence only when legitimate mail is impeded.\/}  it is not as friendly,
nor as catchy, but it more accurately reflects the current situation.} A
typical \gls{SMTP} conversation resembles the following (the lines starting
with a three digit number are sent by the server, all other lines are sent
by the client):

@@

\section{Postfix}

@@

Postfix is a \acrlong{MTA}~(\gls{MTA}) with the following design aims (in
order of importance): security, flexibility of configuration, scalability,
and high performance.  It features extensive optional anti-spam
restrictions, allowing an administrator to deploy those restrictions which
they judge suitable for their site's needs, rather than a fixed set chosen
by Postfix's author.  These restrictions can be selectively applied,
combined, and bypassed on a per-client, per-recipient, or per-sender basis,
allowing different levels of stricture and permissiveness.  Postfix
leverages simple lookup tables to support arbitrarily complicated
user-defined sequences of restrictions and exceptions; policy servers
(\sectionref{policy servers}) provide a simple way to write new
restrictions.  Administrators can also supply their own rejection messages
to make it clear to senders exactly why their mail was rejected.
Unfortunately this flexibility has a cost: complexity in the log files
generated.  Although it is simple to use standard Unix text processing
utilities to determine the fate of an individual email, following the
journey an email takes through Postfix can be quite difficult.  For most
mails the journey is simple and brief, but the remaining minority can have
quite complex journeys (see \sectionref{complications} for details).

@@

\subsection{Mixing and matching Postfix restrictions}

@@

\subsection{Policy servers}

@@

    \item Greylisting~\cite{greylisting} is a technique that temporarily
        rejects a delivery attempt when the triple of \newline{}
        \tab{}\texttt{(sender address, recipient address, remote \gls{IP}
        address)}\newline{} has not been seen before; on second and
        subsequent delivery attempts from that triple the mail will be
        accepted.  This blocks spam because maintaining a list of failed
        addresses and retrying after a temporary failure is uneconomical
        for a spam sender, but a legitimate mail server must retry
        deliveries that temporarily failed.  Sadly spam senders are using
        increasingly complex and well written programs to distribute spam,
        frequently using an ISP provided \gls{SMTP} server from a
        compromised machine on the ISP's network.  Greylisting will slowly
        become less useful, but it does block a large percentage of spam
        mail at the moment; the most effective restrictions from the
        \numberOFlogFILES{} log files used in testing the parser are shown
        in \tableref{Summary of rejections}.  Greylisting is obviously
        worth using, at least at the moment, particularly when you take
        into account Greylisting's position as the final restriction that a
        mail must overcome:\footnote{Greylisting is the final restriction a
        mail must overcome in the configuration used on the mail server the
        log files were obtained from; an administrator is free to use
        Greylisting wherever in the restriction list they feel is most
        appropriate for their mail system.} Greylisting only takes effect
        for mails that have passed every other restriction.  Some problems
        may be encountered when using greylisting: some servers fail to
        retry; legitimate mail may be delayed, particularly when coming
        from a pool of servers.

### 2009-01-16 doc/logparser-conclusion.tex
Minor cleanups.
Converted from article to report, and sections to chapters. Some cleanups. @ doesn't need escaping in glossary entries with new glossaries package.
Lots of minor improvements and cleanups to the State of the Art Review,     including running diction on it. Checked that I've written 'log files' and 'log lines' everywhere, instead of     'files' or 'lines', and added a make target to help with it.  This is     supported by paragraphs-to-lines, which will hopefully find other uses.  I     renamed a couple of included files too, to avoid false positives. Improved the check for cont

\section{Conclusion}

@@

Given the diverse nature of the log lines and the ease with which
administrators can cause new lines to be logged (\sectionref{postfix
background}), enabling users to easily extend the parser to deal with new
log lines is a design imperative (\sectionref{parser design}).  Despite the
resulting initial increase in complexity the task is quite tractable,
though it does raise efficiency and optimisation questions (answered in
\sectionref{parser efficiency}).  Providing a tool to ease the generation
of regexes from unparsed log lines should greatly help users add rules
to parse previously unparsed log lines (\sectionref{creating new rules}).

@@

% XXX THIS SHOULD PROBABLY BE REMOVED\@.
%The emergent behaviour (\sectionref{Emergent
%behaviour}) exhibited by the rules and actions is also interesting, and is
%discussed after the flow chart (\sectionref{flow-chart}) and explanation of
%the paths through the parser.  This emergent behaviour greatly eases the
%process of adding new actions (\sectionref{adding new actions}), as the
%actions do not have to be inserted into an explicit flow of control; new
%actions will naturally find their place in the paths through the parser.

@@

The real difficulties arise once the parser is successfully dealing with
90\% of the log lines, as the irregularities and complications explained in
\sectionref{complications} only become apparent once the vast majority of
log lines have been parsed successfully.  Adding new rules to deal with
numerous, infrequently occurring log lines is a simple, albeit tiresome,
task, (though alleviated by the tool described in \sectionref{adding new
actions}), whereas dealing with mails where information appears to be
missing is much more grueling.  Trawling through the log files, looking for
something out of the ordinary, possibly hundreds or even thousands of lines
away from the last mention of the queueid in question, is extremely time
consuming and error prone.  Sometimes the task is not to spot the line that
is unusual, but to spot that a line that is normally present is missing,
i.e.\ to realise that one line amongst thousands is absent.  In all cases
the evidence must be used to construct a hypothesis to explain the
irregularities, and that hypothesis must then be tested in the parser; if
successful, the parser must be modified to deal with the irregularities,
without adversely affecting the existing functionality.  The complications
described in \sectionref{complications} were solved in the order they are
described in, and that order closely resembles the frequency with which
they occur; the most frequently occurring complications dominate the
warning messages produced, and so naturally they are the first
complications to be dealt with.

>>>>

\chapter{Conclusion}

@@

Given the diverse nature of the log lines and the ease with which
administrators can cause new log lines to be logged (\sectionref{postfix
background}), enabling users to easily extend the parser to deal with new
log lines is a design imperative (\sectionref{parser design}).  Despite the
resulting initial increase in complexity the task is quite tractable,
though it does raise efficiency and optimisation questions (answered in
\sectionref{parser efficiency}).  Providing a tool to ease the generation
of regexes from unparsed log lines should greatly help users add rules
to parse previously unparsed log lines (\sectionref{creating new rules}).

@@

% XXX THIS SHOULD PROBABLY BE REMOVED\@.
%The emergent behaviour (\sectionref{Emergent behaviour}) exhibited by the
%rules and actions is also interesting, and is discussed after the flow
%chart (\sectionref{flow-chart}) and explanation of the paths through the
%parser.  This emergent behaviour greatly eases the process of adding new
%actions (\sectionref{adding new actions}), as the actions do not have to
%be inserted into an explicit flow of control; new actions will naturally
%find their place in the paths through the parser.

@@

The real difficulties arise once the parser is successfully dealing with
90\% of the log lines, as the irregularities and complications explained in
\sectionref{complications} only become apparent once the vast majority of
log lines have been parsed successfully.  Adding new rules to deal with
numerous, infrequently occurring log lines is a simple, albeit tiresome,
task, (though alleviated by the tool described in \sectionref{adding new
actions}), whereas dealing with mails where information appears to be
missing is much more grueling.  Trawling through the log files, looking for
something out of the ordinary, possibly hundreds or even thousands of log
lines away from the last mention of the queueid in question, is extremely
time consuming and error prone.  Sometimes the task is not to spot the line
that is unusual, but to spot that a line that is normally present is
missing, i.e.\ to realise that one line amongst thousands is absent.  In
all cases the evidence must be used to construct a hypothesis to explain
the irregularities, and that hypothesis must then be tested in the parser;
if successful, the parser must be modified to deal with the irregularities,
without adversely affecting the existing functionality.  The complications
described in \sectionref{complications} were solved in the order they are
described in, and that order closely resembles the frequency with which
they occur; the most frequently occurring complications dominate the
warning messages produced, and so naturally they are the first
complications to be dealt with.

### 2009-01-16 doc/logparser-implementation.tex
Improvements to the glossaries appearance. Minor fixes.
Converted from article to report, and sections to chapters. Some cleanups. @ doesn't need escaping in glossary entries with new glossaries package.
Lots of minor improvements and cleanups to the State of the Art Review,     including running diction on it. Checked that I've written 'log files' and 'log lines' everywhere, instead of     'files' or 'lines', and added a make target to help with it.  This is     supported by paragraphs-to-lines, which will hopefully find other uses.  I     renamed a couple of included files too, to avoid fals

\section{Postfix Parser Implementation}
\label{Postfix Parser Implementation}

@@

\subsection{Assumptions}

@@

\subsection{Database Schema}
\label{database schema}

@@

\subsubsection{Rules table}

@@

\subsubsection{Connections table}

@@

\subsubsection{Results table}

@@

\subsection{Rules}

@@

\subsubsection{Example rule}

@@

\subsubsection{Creating new rules}

@@

    \item [Calculate word frequencies]  The position of words within a log
        line is important: a word occurring in two lines does not indicate
        similarity unless it occupies the same position within both log
        lines.\footnote{If a variable term within a line contains spaces,
        it will appear to the algorithm as two or more words rather than
        one.  This will alter the position of subsequent words, so a word
        occurring in different positions in two log lines \textit{may\/}
        indicate similarity, but the algorithm does not attempt to deal
        with this possibility.}  The algorithm maintains a counter for each
        \texttt{(word, word's position within the log line)} tuple,
        incrementing it each time that word occurs in that position.

@@

\subsubsection{Actions in detail}

@@

    \item [BOUNCE] Postfix 2.3 and subsequent versions log the creation of
        bounce messages.  This action creates a new mail if necessary; if
        the mail already exists the unknown origin flag will be removed.
        The action also marks the mail as a bounce notification.  To deal
        with complication \sectionref{Bounce notification mails delivered
        before their creation is logged} this action checks a cache of
        recent bounce mails to avoid creating bogus bounce mails when lines
        are logged out of order.

@@

    \item [MAIL\_QUEUED] This action represents Postfix picking a mail from
        the queue to deliver. This action is used for both \daemon{qmgr}
        and \daemon{cleanup} as it needs to deal with out of order log
        lines; see \sectionref{discarding cleanup lines} for details.
        XXXXXX EXPLAIN THIS BETTER\@.

@@

\subsubsection{Adding new actions}

@@

\subsection{Framework}

@@

\subsection{Parser Flow Chart}

@@

\subsubsection{Mail enters the system}

@@

\subsubsection{Mail delivery}

@@

\subsubsection{Emergent behaviour}

@@

XXX DUNNO IF I SHOULD INCLUDE THIS SECTION\@; AT BEST IT NEEDS TO BE
UPDATED, AT WORST IT'S RUBBISH\@.

@@

\subsection{Complications encountered}

@@

\subsubsection{Queueid vs pid}

@@

\subsubsection{Connection reuse}

@@

\subsubsection{Re-injected mails}

@@

Because of the uncertain order the parser cannot complain when it
encounters a log line from \daemon{qmgr} for a previously unseen mail;
instead it must flag the mail as coming from an unknown origin, and
subsequently clear the flag if and when the origin of the mail becomes
clear.  Obviously the parser could omit checking where mails originate
from, but requiring an explicit source helps to expose bugs in the parser;
such checks helped to identify the complications described in
\sectionref{discarding cleanup lines} and \sectionref{pickup logging after
cleanup}.

@@

\subsubsection{Identifying bounce notifications}

@@

\subsubsection{Aborted delivery attempts}

@@

    \item If there are third and subsequent results, are all their
        postfix\_action attributes equal to INFO\@?  If there are any log
        lines after the first two they should be informational lines only.

@@

\subsubsection{Further aborted delivery attempts}

@@

    \item Were there three or more \daemon{smtpd} log lines for the mail?
        There should be a connection line and a mail accepted line,
        followed by one or more rejection lines.

@@

\subsubsection{Timeouts during DATA phase}

@@

XXX MAKE THIS PARAGRAPH CLEARER\@.  A timeout may thus apply either to an
accepted mail or a rejected mail.  To distinguish between the two cases the
algorithm compares the timestamp of the last accepted mail against the
timestamp of the last line logged by \daemon{smtpd} for that connection
(the TIMEOUT action is dependant on the CLONE action keeping a list of all
mails accepted on each connection).  If the mail acceptance timestamp is
later then the timeout applies to the just-accepted mail, which will be
discarded.  If the \daemon{smtpd} timestamp is later there was a rejection
between the accepted mail and the timeout: the action assumes that the
timeout applies to a rejected delivery attempt and finishes.  This
assumption is not necessarily correct, because Postfix may have accepted an
earlier recipient and rejected a later one, in which case the timeout
applies to the accepted mail, which should be discarded.  This has not been
a problem in practice, though it may be in future.  This complication is
further complicated by the presence of out of order \daemon{cleanup} lines:
see \sectionref{discarding cleanup lines} for details.

@@

\subsubsection{Discarding cleanup lines}

\label{discarding cleanup lines}

@@

\subsubsection{Pickup logging after cleanup}

@@

When mail is submitted locally, \daemon{pickup} accepts the new mail and
generates a log line showing the source.  Occasionally this log line will
occur later in the log file than the \daemon{cleanup} log line, so the
PICKUP action will find that a state table entry exists for that queueid.
Normally if the queueid given in the PICKUP line exists in the state tables
a warning is generated by the \daemon{pickup} action, but if the following
conditions are met it is assumed that the lines were out of order:

@@

    \item There is less than a five second difference between the
        timestamps of the \daemon{cleanup} and \daemon{pickup} lines.

@@

\subsubsection{Smtpd stops logging}

@@

\subsubsection{Out of order log lines}

@@

Occasionally a log file will have out of order log lines which cannot be
dealt with by the techniques described in \sectionref{tracking re-injected
mail}, \sectionref{discarding cleanup lines} or \sectionref{pickup logging
after cleanup}.  In the \numberOFlogFILES{} log files used for testing this
occurs only five times in 60,721,709 log lines, but for parser correctness
it must be dealt with.  The five occurrences have the same characteristics:
the \daemon{local} log line showing delivery to a local mailbox occurs
after the \daemon{qmgr} log line showing removal of the mail from the queue
because delivery is completed.  This causes problems: the data in the state
tables for the mail is not complete, so entering it into the database
fails; a new mail is created when the \daemon{local} line is parsed and
remains in the state tables; four warnings are issued per pair of out of
order log lines.

The COMMIT action examines the list of programs that have logged lines for
each mail, comparing the list against a table of known-good program
combinations.  If the mail's combination is found in the table the mail can
be entered in the database; if the combination is not found entry must be
postponed and the mail flagged for later entry.  The SAVE\_DATA action
checks for that flag; if the additional log lines have caused the mail to
reach a valid combination then entry in the database will proceed,
otherwise it must be postponed once more.

@@

\subsubsection{Yet more aborted delivery attempts}

@@

    \item IS THIS CORRECT NOW\@?  CAN THERE BE MORE smtpd LINES\@?  There
        must be exactly two \daemon{smtpd} and one \daemon{cleanup} log
        lines for the mail, with no additional log lines.

@@

\subsubsection{Mail deleted before delivery is attempted}

@@

\subsubsection{Bounce notification mails delivered before their creation is
logged}

@@

\subsubsection{Mails deleted during delivery}

@@

\subsection{Limitations and possible improvements}

@@

\subsubsection{Limitations}

@@

\subsubsection{Possible improvements}

>>>>

\chapter{Postfix Parser Implementation}
\label{Postfix Parser Implementation}

@@

\section{Assumptions}

@@

\section{Database Schema}
\label{database schema}

@@

\subsection{Rules table}

@@

\subsection{Connections table}

@@

\subsection{Results table}

@@

\section{Rules}

@@

\subsection{Example rule}

@@

\subsection{Creating new rules}

@@

    \item [Calculate word frequencies]  The position of words within a log
        line is important: a word occurring in two log lines does not
        indicate similarity unless it occupies the same position within
        both log lines.\footnote{If a variable term within a line contains
        spaces, it will appear to the algorithm as two or more words rather
        than one.  This will alter the position of subsequent words, so a
        word occurring in different positions in two log lines
        \textit{may\/} indicate similarity, but the algorithm does not
        attempt to deal with this possibility.}  The algorithm maintains a
        counter for each \texttt{(word, word's position within the log
        line)} tuple, incrementing it each time that word occurs in that
        position.

@@

\subsection{Actions in detail}

@@

    \item [BOUNCE] Postfix 2.3 and subsequent versions log the creation of
        bounce messages.  This action creates a new mail if necessary; if
        the mail already exists the unknown origin flag will be removed.
        The action also marks the mail as a bounce notification.  To deal
        with complication \sectionref{Bounce notification mails delivered
        before their creation is logged} this action checks a cache of
        recent bounce mails to avoid creating bogus bounce mails when log
        lines are out of order.

@@

    \item [MAIL\_QUEUED] This action represents Postfix picking a mail from
        the queue to deliver. This action is used for both \daemon{qmgr}
        and \daemon{cleanup} as it needs to deal with out of order log
        lines; see \sectionref{discarding cleanup log lines} for details.
        XXX EXPLAIN THIS BETTER\@.

@@

\subsection{Adding new actions}

@@

\section{Framework}

@@

\section{Parser Flow Chart}

@@

\subsection{Mail enters the system}

@@

\subsection{Mail delivery}

@@

\subsection{Emergent behaviour}

@@

XXX DUNNO IF I SHOULD INCLUDE THIS SECTION\@; AT BEST IT NEEDS TO BE
UPDATED, AT WORST IT IS RUBBISH\@.

@@

\section{Complications encountered}

@@

\subsection{Queueid vs pid}

@@

\subsection{Connection reuse}

@@

\subsection{Re-injected mails}

@@

Because of the uncertain order the parser cannot complain when it
encounters a log line from \daemon{qmgr} for a previously unseen mail;
instead it must flag the mail as coming from an unknown origin, and
subsequently clear the flag if and when the origin of the mail becomes
clear.  Obviously the parser could omit checking where mails originate
from, but requiring an explicit source helps to expose bugs in the parser;
such checks helped to identify the complications described in
\sectionref{discarding cleanup log lines} and \sectionref{pickup logging
after cleanup}.

@@

\subsection{Identifying bounce notifications}

@@

\subsection{Aborted delivery attempts}

@@

    \item If there are third and subsequent results, are all their
        postfix\_action attributes equal to INFO\@?  If there are any log
        lines after the first two they should be informational only.

@@

\subsection{Further aborted delivery attempts}

@@

    \item Were there three or more \daemon{smtpd} log lines for the mail?
        There should be a connection log line and a mail accepted log line,
        followed by one or more rejection log lines.

@@

\subsection{Timeouts during DATA phase}

@@

XXX MAKE THIS PARAGRAPH CLEARER\@.  A timeout may thus apply either to an
accepted mail or a rejected mail.  To distinguish between the two cases the
algorithm compares the timestamp of the last accepted mail against the
timestamp of the last line logged by \daemon{smtpd} for that connection
(the TIMEOUT action is dependant on the CLONE action keeping a list of all
mails accepted on each connection).  If the mail acceptance timestamp is
later then the timeout applies to the just-accepted mail, which will be
discarded.  If the \daemon{smtpd} timestamp is later there was a rejection
between the accepted mail and the timeout: the action assumes that the
timeout applies to a rejected delivery attempt and finishes.  This
assumption is not necessarily correct, because Postfix may have accepted an
earlier recipient and rejected a later one, in which case the timeout
applies to the accepted mail, which should be discarded.  This has not been
a problem in practice, though it may be in future.  This complication is
further complicated by the presence of out of order \daemon{cleanup} log
lines: see \sectionref{discarding cleanup log lines} for details.

@@

\subsection{Discarding cleanup log lines}

\label{discarding cleanup log lines}

@@

\subsection{Pickup logging after cleanup}

@@

When mail is submitted locally, \daemon{pickup} accepts the new mail and
generates a log line showing the source.  Occasionally this log line will
occur later in the log file than the \daemon{cleanup} log line, so the
PICKUP action will find that a state table entry exists for that queueid.
Normally if the queueid given in the PICKUP line exists in the state tables
a warning is generated by the \daemon{pickup} action, but if the following
conditions are met it is assumed that the log lines were out of order:

@@

    \item There is less than a five second difference between the
        timestamps of the \daemon{cleanup} and \daemon{pickup} log lines.

@@

\subsection{Smtpd stops logging}

@@

\subsection{Out of order log lines}

@@

Occasionally a log file will have out of order log lines which cannot be
dealt with by the techniques described in \sectionref{tracking re-injected
mail}, \sectionref{discarding cleanup log lines} or \sectionref{pickup
logging after cleanup}.  In the \numberOFlogFILES{} log files used for
testing this occurs only five times in 60,721,709 log lines, but for parser
correctness it must be dealt with.  The five occurrences have the same
characteristics: the \daemon{local} log line showing delivery to a local
mailbox occurs after the \daemon{qmgr} log line showing removal of the mail
from the queue because delivery is completed.  This causes problems: the
data in the state tables for the mail is not complete, so entering it into
the database fails; a new mail is created when the \daemon{local} line is
parsed and remains in the state tables; four warnings are issued per pair
of out of order log lines.

The COMMIT action examines the list of programs that have produced a log
lines for each mail, comparing the list against a table of known-good
program combinations.  If the mail's combination is found in the table the
mail can be entered in the database; if the combination is not found entry
must be postponed and the mail flagged for later entry.  The SAVE\_DATA
action checks for that flag; if the additional log lines have caused the
mail to reach a valid combination then entry in the database will proceed,
otherwise it must be postponed once more.

@@

\subsection{Yet more aborted delivery attempts}

@@

    \item XXX IS THIS CORRECT NOW\@?  CAN THERE BE MORE smtpd LOG LINES\@?
        There must be exactly two \daemon{smtpd} and one \daemon{cleanup}
        log lines for the mail, with no additional log lines.

@@

\subsection{Mail deleted before delivery is attempted}

@@

\subsection{Bounce notification mails delivered before their creation is
logged}

@@

\subsection{Mails deleted during delivery}

@@

\section{Limitations and possible improvements}

@@

\subsection{Limitations}

@@

\subsection{Possible improvements}
